---
title: "Deprivation, Access, and Food Purchasing in London: A Multi-Stage Analysis"
author: "[Author Name]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    fig_width: 10
    fig_height: 7
    theme: cosmo
    highlight: tango
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache   = TRUE
)

# ── Packages ──────────────────────────────────────────────────────────
required_packages <- c("sf", "dplyr", "ggplot2", "patchwork", "scales",
                       "tidyr", "spdep", "cluster", "factoextra")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org", quiet = TRUE)
  }
}

library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)
library(scales)
library(tidyr)

sf_use_s2(FALSE)  # Required for projected data (EPSG:27700)
```

```{r theme-setup}
# ── Color Palettes (Okabe-Ito, Nature-Recommended) ──────────────────────
# Categorical palette (up to 8 categories)
pal_categorical <- c(
  "#E69F00",  # Orange — primary accent
  "#56B4E9",  # Sky Blue — secondary accent
  "#009E73",  # Bluish Green
  "#F0E442",  # Yellow (use sparingly, poor contrast on white)
  "#CC79A7",  # Pink/Violet
  "#0077BB",  # Blue
  "#D55E00",  # Vermilion
  "#000000"   # Black
)

# Store type colors (3 categories — from categorical, picking best contrast)
pal_store_type <- c(
  "Express"    = "#0077BB",  # Blue
  "Superstore" = "#009E73",  # Green
  "Extra"      = "#D55E00"   # Vermilion/Orange-Red
)

# Model comparison colors (4 methods)
pal_model <- c(
  "OLS"      = "#CC79A7",  # Pink
  "LASSO"    = "#0077BB",  # Blue
  "XGBoost"  = "#009E73",  # Green
  "IMD Total" = "#E69F00"  # Orange
)

# Cluster colors (4 clusters — distinct, colorblind-safe)
pal_cluster <- c(
  "Standard Urban"       = "#56B4E9",  # Sky Blue
  "Superstore-dominant"  = "#E69F00",  # Orange
  "Full Service"         = "#009E73",  # Green
  "Underserved"          = "#D55E00"   # Vermilion
)

# SHAP gradient (blue-white-red)
shap_gradient <- c(low = "#2166AC", mid = "#F7F7F7", high = "#B2182B")

# Accent for LOESS/smooth lines
accent_line <- "#CC3311"

# ── Unified Base Theme ──────────────────────────────────────────────────
theme_pub <- function(base_size = 11, map_mode = FALSE) {
  ret <- theme_minimal(base_size = base_size) +
    theme(
      # Typography
      plot.title    = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(size = base_size - 1, colour = "grey40"),
      plot.caption  = element_text(size = base_size - 3, colour = "grey50",
                                   hjust = 1, margin = margin(t = 4)),
      strip.text    = element_text(face = "bold", size = base_size),

      # Legend
      legend.position    = "bottom",
      legend.text        = element_text(size = base_size - 2),
      legend.title       = element_text(face = "bold", size = base_size - 1),
      legend.key.width   = unit(1.2, "cm"),
      legend.margin      = margin(4, 4, 4, 4),
      legend.box.margin  = margin(0, 0, 4, 0),

      # Grid
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),

      # Axes
      axis.title  = element_text(size = base_size),
      axis.text   = element_text(size = base_size - 1),

      # Margins
      plot.margin = margin(8, 8, 8, 8, unit = "pt")
    )

  if (map_mode) {
    ret <- ret + theme(
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
  }

  ret
}
```

```{r paths}
# Resolve project root (two levels up from scripts/)
PROJ <- normalizePath("..")
CSV  <- file.path(PROJ, "output/csv")
SHP  <- file.path(PROJ, "output/shp")
```

---

## Introduction

The conventional food desert paradigm assumes that limited access to affordable, nutritious food explains poor dietary outcomes. However, this supply-side view neglects the complex interplay between geography, deprivation, and actual purchasing behavior. Our research addresses this gap by examining how area-level deprivation and its sub-domains—when compared to supply-side accessibility metrics and socio-demographic factors—explain spatial variation in food purchasing patterns across London.

Building on recent work by Broadbridge et al. (2025), who demonstrated that store proximity explains far less about purchasing healthiness than socio-demographic factors, and Beydoun et al. (2026), who revealed that aggregate deprivation indices mask critical sub-domain heterogeneity, we employ a three-stage analytical approach. First, we characterize spatial heterogeneity in Tesco store accessibility across London. Second, we decompose the Index of Multiple Deprivation (IMD) into its seven constituent domains to test which dimensions best predict food purchasing patterns. Third, we integrate supply-side metrics, deprivation sub-domains, and socio-demographic controls in a non-linear machine learning framework to identify which factors retain predictive power when competing directly.

---

## 1. Import POI Data

```{r poi}
poi <- read.csv(file.path(CSV, "poi_tesco_stores.csv"))

# Remove non-food stores (Tesco Mobile)
poi <- poi %>% filter(store_type != "Unknown")

cat(sprintf("Loaded %d Tesco stores\n", nrow(poi)))
cat("Store type distribution:\n")
print(table(poi$store_type))

# Convert to sf (WGS84 → British National Grid)
poi_sf <- st_as_sf(poi, coords = c("lon", "lat"), crs = "EPSG:4326")
poi_sf <- st_transform(poi_sf, crs = 27700)

cat(sprintf("\nCRS: EPSG:%d\n", st_crs(poi_sf)$epsg))
```

---

## 2. Generate 1250m Walking Buffers

A 15-minute walking distance (approximately 1.25 km at 5 km/h) represents a reasonable threshold for foot-based food access in urban environments. For each store, we generate circular buffers representing this accessible area, then union buffers by store type to create continuous service areas—avoiding boundary artifacts that plague raster-based approaches.

```{r buffers}
BUFFER_DIST <- 1250  # metres

# Create buffer for each store
poi_sf <- poi_sf %>%
  mutate(buffer = st_buffer(geometry, dist = BUFFER_DIST))

# Union buffers by store type (3 polygons instead of 565 — critical for performance)
buffers_by_type <- poi_sf %>%
  group_by(store_type) %>%
  summarise(geometry = st_union(buffer), .groups = "drop")

# Union of ALL buffers (for total_coverage)
buffer_all <- poi_sf %>%
  summarise(geometry = st_union(buffer))

cat(sprintf("Buffer union by type: %d rows\n", nrow(buffers_by_type)))
cat(sprintf("Store types: %s\n", paste(buffers_by_type$store_type, collapse = ", ")))
```

```{r boundaries}
# Load shapefiles (geometry only, strip pre-computed columns)
lsoa_geom <- st_read(file.path(SHP, "lsoa_boundaries/lsoa_boundaries.shp"),
                     quiet = TRUE) %>%
  select(lsoa_code, geometry) %>%
  st_make_valid()

msoa_geom <- st_read(file.path(SHP, "msoa_boundaries/msoa_boundaries.shp"),
                     quiet = TRUE) %>%
  select(MSOA11CD, geometry) %>%
  st_make_valid()

ward_geom <- st_read(file.path(SHP, "ward_boundaries/ward_boundaries.shp"),
                     quiet = TRUE) %>%
  select(WD11CD, geometry) %>%
  st_make_valid()

bor_geom <- st_read(file.path(SHP, "borough_boundaries/borough_boundaries.shp"),
                    quiet = TRUE) %>%
  select(bor_code, geometry) %>%
  rename(borough_code = bor_code) %>%
  st_make_valid()

# Verify CRS
stopifnot(st_crs(lsoa_geom)$epsg == 27700)

cat("Boundary data loaded:\n")
cat(sprintf("  LSOA:    %d units\n", nrow(lsoa_geom)))
cat(sprintf("  MSOA:    %d units\n", nrow(msoa_geom)))
cat(sprintf("  Ward:    %d units\n", nrow(ward_geom)))
cat(sprintf("  Borough: %d units\n", nrow(bor_geom)))
```

## 3. Visualize Buffers Over London

```{r buffer-viz, fig.width = 10, fig.height = 8}
ggplot() +
  geom_sf(data = msoa_geom, fill = NA, colour = "grey70", linewidth = 0.1) +
  geom_sf(data = buffers_by_type, aes(fill = store_type), alpha = 0.3, colour = NA) +
  geom_sf(data = poi_sf, aes(colour = store_type), size = 0.8, alpha = 0.6) +
  scale_fill_manual(values = pal_store_type, name = "Store Type") +
  scale_colour_manual(values = pal_store_type, name = "Store Type") +
  labs(
    title = "Tesco Store Locations with 1250m Walking Buffers",
    subtitle = sprintf("%d stores | Buffer = 15-min walk (%dm)", nrow(poi_sf), BUFFER_DIST),
    caption = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | Boundaries: ONS Open Geography"
  ) +
  theme_pub(base_size = 11, map_mode = TRUE) +
  theme(legend.position = "right")
```

## 5. Coverage Computation Function

Coverage is computed independently at each geographic scale by intersecting
the unioned buffer polygons with boundary polygons.

**Formula:** `coverage% = area(boundary ∩ buffer) / area(boundary) × 100`

This normalizes for geographic unit size — larger units don't get inflated coverage.

```{r coverage-function}
compute_coverage <- function(boundaries, buffers_by_type, buffer_all, id_col) {

  # Compute area of each boundary unit (projected CRS → square metres)
  boundaries <- boundaries %>%
    mutate(area_m2 = as.numeric(st_area(geometry)))

  id_col_sym <- sym(id_col)
  store_types <- buffers_by_type$store_type
  cov_cols <- character(0)

  for (st in store_types) {
    buf <- buffers_by_type %>% filter(store_type == !!st)
    col_name <- paste0(tolower(st), "_coverage")

    # st_intersection only returns rows WHERE intersection exists
    inter <- st_intersection(boundaries %>% select(all_of(id_col), geometry), buf)

    # Sum overlap areas per geographic unit
    overlap_df <- inter %>%
      mutate(overlap_m2 = as.numeric(st_area(geometry))) %>%
      as.data.frame() %>%
      group_by(!!id_col_sym) %>%
      summarise(overlap_m2 = sum(overlap_m2), .groups = "drop") %>%
      rename(!!col_name := overlap_m2)

    # Join back — units with 0% coverage get NA, then 0
    boundaries <- boundaries %>%
      left_join(overlap_df, by = id_col) %>%
      mutate(!!col_name := (!!sym(col_name)) / area_m2 * 100,
             !!col_name := replace_na(!!sym(col_name), 0))

    cov_cols <- c(cov_cols, col_name)
  }

  # Total coverage (union of ALL store types)
  inter_all <- st_intersection(
    boundaries %>% select(all_of(id_col), geometry),
    buffer_all
  )
  total_df <- inter_all %>%
    mutate(total_m2 = as.numeric(st_area(geometry))) %>%
    as.data.frame() %>%
    group_by(!!id_col_sym) %>%
    summarise(total_m2 = sum(total_m2), .groups = "drop") %>%
    rename(total_coverage = total_m2)

  boundaries <- boundaries %>%
    left_join(total_df, by = id_col) %>%
    mutate(
      total_coverage = total_coverage / area_m2 * 100,
      total_coverage = replace_na(total_coverage, 0)
    )

  # Derived metrics
  boundaries <- boundaries %>%
    mutate(
      overlap_index = !!sym(cov_cols[1]) + !!sym(cov_cols[2]) + !!sym(cov_cols[3]) - total_coverage,
      unique_types = (!!sym(cov_cols[1]) > 0) + (!!sym(cov_cols[2]) > 0) + (!!sym(cov_cols[3]) > 0)
    )

  # Clean up temp column
  boundaries <- boundaries %>% select(-area_m2)

  return(boundaries)
}
```

---

## 6. Compute Coverage at All 4 Scales

```{r compute-coverage, cache.extra = 1}
cat("Computing LSOA coverage ...\n")
lsoa_cov <- compute_coverage(lsoa_geom, buffers_by_type, buffer_all, "lsoa_code")
cat(sprintf("  Done: %d rows\n", nrow(lsoa_cov)))

cat("Computing MSOA coverage ...\n")
msoa_cov <- compute_coverage(msoa_geom, buffers_by_type, buffer_all, "MSOA11CD")
cat(sprintf("  Done: %d rows\n", nrow(msoa_cov)))

cat("Computing Ward coverage ...\n")
ward_cov <- compute_coverage(ward_geom, buffers_by_type, buffer_all, "WD11CD")
cat(sprintf("  Done: %d rows\n", nrow(ward_cov)))

cat("Computing Borough coverage ...\n")
bor_cov <- compute_coverage(bor_geom, buffers_by_type, buffer_all, "borough_code")
cat(sprintf("  Done: %d rows\n", nrow(bor_cov)))
```

---

## 7. Verify & Summarize Coverage

```{r check-na}
cov_cols <- c("express_coverage", "superstore_coverage", "extra_coverage",
              "total_coverage", "overlap_index", "unique_types")

for (name in c("lsoa_cov", "msoa_cov", "ward_cov", "bor_cov")) {
  df <- get(name)
  na_counts <- sapply(cov_cols, function(c) sum(is.na(df[[c]])))
  total_na  <- sum(na_counts)
  cat(sprintf("%s (%d rows): %d NAs total", name, nrow(df), total_na))
  if (total_na > 0) {
    cat(" ← ")
    cat(paste(names(na_counts[na_counts > 0]), collapse = ", "))
  }
  cat("\n")
}

# Out-of-bounds: values > 100 from floating point precision in st_intersection
for (name in c("lsoa_cov", "msoa_cov", "ward_cov", "bor_cov")) {
  df <- get(name)
  for (c in c("express_coverage", "superstore_coverage", "extra_coverage", "total_coverage")) {
    n_above <- sum(df[[c]] > 100, na.rm = TRUE)
    if (n_above > 0) cat(sprintf("  %s.%s: %d values > 100\n", name, c, n_above))
  }
}
```

```{r shapefile-check, fig.width = 12, fig.height = 10}
plot_boundary <- function(gdf, title) {
  ggplot(gdf) +
    geom_sf(fill = "#2ca02c", colour = "white", linewidth = 0.1) +
    labs(title = title) +
    theme_pub(base_size = 10, map_mode = TRUE) +
    theme(plot.title = element_text(face = "bold"))
}

plot_area <- function(gdf, title) {
  gdf_area <- gdf %>% mutate(area_m2 = as.numeric(st_area(geometry)))
  ggplot(gdf_area) +
    geom_sf(aes(fill = area_m2), colour = "white", linewidth = 0.1) +
    scale_fill_viridis_c(option = "plasma", na.value = "grey50") +
    labs(title = title) +
    theme_pub(base_size = 10, map_mode = TRUE) +
    theme(plot.title = element_text(face = "bold"))
}

(plot_boundary(lsoa_geom, "LSOA") | plot_boundary(msoa_geom, "MSOA")) /
  (plot_boundary(ward_geom, "Ward") | plot_boundary(bor_geom, "Borough")) +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(
    title = "Shapefile & Area Check",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, colour = "grey40")
    )
  )

(plot_area(lsoa_geom, "LSOA area_m2") | plot_area(msoa_geom, "MSOA area_m2")) /
  (plot_area(ward_geom, "Ward area_m2") | plot_area(bor_geom, "Borough area_m2")) +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(
    title = "Area Check (viridis plasma)",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, colour = "grey40")
    )
  )
```

```{r summary-table}
cov_metrics <- c("express_coverage", "superstore_coverage", "extra_coverage",
                 "total_coverage", "overlap_index", "unique_types")

make_summary <- function(df, scale_name) {
  data.frame(
    Scale = scale_name,
    n = nrow(df),
    Metric = cov_metrics,
    Mean = sapply(cov_metrics, function(m) mean(df[[m]], na.rm = TRUE)),
    SD = sapply(cov_metrics, function(m) sd(df[[m]], na.rm = TRUE)),
    Min = sapply(cov_metrics, function(m) min(df[[m]], na.rm = TRUE)),
    Median = sapply(cov_metrics, function(m) median(df[[m]], na.rm = TRUE)),
    Max = sapply(cov_metrics, function(m) max(df[[m]], na.rm = TRUE))
  )
}

summary_all <- bind_rows(
  make_summary(lsoa_cov, "LSOA"),
  make_summary(msoa_cov, "MSOA"),
  make_summary(ward_cov, "Ward"),
  make_summary(bor_cov, "Borough")
)

knitr::kable(summary_all, digits = 1,
             caption = "Coverage Summary Across Geographic Scales")
```


---

## 8. Multi-Scale Comparison (2×2 Panel)

```{r helper}
make_panel <- function(gdf, title, point_size = 0.3) {
  p <- ggplot(gdf) +
    geom_sf(aes(fill = total_coverage), colour = NA, linewidth = 0) +
    geom_sf(data = poi_sf, colour = "black", size = point_size, alpha = 0.4) +
    scale_fill_viridis_c(
      option = "plasma", begin = 0.1, end = 0.95,
      breaks = seq(0, 100, by = 20),
      labels = percent_format(scale = 1, suffix = "%")
    ) +
    labs(
      title = title,
      subtitle = sprintf("n = %d | Mean = %.1f%% | SD = %.1f%%",
                         nrow(gdf),
                         mean(gdf$total_coverage, na.rm = TRUE),
                         sd(gdf$total_coverage, na.rm = TRUE))
    ) +
    theme_pub(base_size = 10, map_mode = TRUE) +
    theme(
      legend.position  = "none",
      plot.title       = element_text(face = "bold", size = 12),
      plot.subtitle    = element_text(size = 8, colour = "grey40")
    )
  # Boundary lines for coarser geographies
  if (nrow(gdf) <= 40) {
    p <- p + geom_sf(colour = "white", linewidth = 0.3, fill = NA)
  }
  p
}
```

```{r multiscale, fig.width = 12, fig.height = 10}
p1 <- make_panel(lsoa_cov, "LSOA Level",           point_size = 0.2)
p2 <- make_panel(msoa_cov, "MSOA Level",           point_size = 0.3)
p3 <- make_panel(ward_cov, "Ward Level",           point_size = 0.3)
p4 <- make_panel(bor_cov,  "London Borough Level", point_size = 0.8)

# Shared legend on MSOA panel
p2 <- p2 +
  theme(legend.position  = "right",
        legend.key.width = unit(1.5, "cm")) +
  labs(fill = "Coverage\n(% area)")

combined <- (p1 | p2) / (p3 | p4) +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(
    title    = "Tesco Store Accessibility Across London: Multi-Scale Comparison",
    subtitle = "15-minute walking buffer coverage (% of area covered by at least one Tesco store)",
    caption  = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | Boundaries: ONS Open Geography | LSOA (~1,500 residents) | MSOA (~7,200) | Ward (electoral division) | Borough (~250,000)",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      plot.caption  = element_text(size = 7, colour = "grey50", hjust = 0.5)
    )
  )

combined
```

The multi-scale panel demonstrates how coverage patterns change with aggregation level. LSOA coverage appears nearly uniform due to small geographic size, while borough coverage shows only coarse differentiation. MSOA coverage reveals the most meaningful spatial variation, confirming our methodological choice for subsequent analysis—balancing sufficient spatial resolution with meaningful differentiation across London.

---

## 9. MSOA Coverage Maps

MSOA is chosen as the analysis scale based on the multi-scale comparison above.

```{r msoa-by-type, fig.width = 14, fig.height = 5}
make_type_panel <- function(gdf, metric, title) {
  ggplot(gdf) +
    geom_sf(aes_string(fill = metric), colour = NA, linewidth = 0) +
    geom_sf(data = poi_sf, colour = "black", size = 0.2, alpha = 0.3) +
    scale_fill_viridis_c(
      option = "plasma", begin = 0.1, end = 0.95,
      breaks = seq(0, 100, by = 20),
      labels = percent_format(scale = 1, suffix = "%")
    ) +
    labs(title = title) +
    theme_pub(base_size = 10, map_mode = TRUE) +
    theme(
      legend.position  = "none",
      plot.title       = element_text(face = "bold")
    )
}

p_exp <- make_type_panel(msoa_cov, "express_coverage",    "Express")
p_sup <- make_type_panel(msoa_cov, "superstore_coverage", "Superstore")
p_ext <- make_type_panel(msoa_cov, "extra_coverage",      "Extra")

p_sup <- p_sup +
  theme(legend.position = "right", legend.key.width = unit(1.5, "cm")) +
  labs(fill = "Coverage\n(% area)")

(p_exp | p_sup | p_ext) +
  plot_annotation(
    title    = "MSOA Coverage by Store Type",
    subtitle = "1250m walking buffer",
    caption  = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | Boundaries: ONS Open Geography",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      plot.caption  = element_text(size = 7, colour = "grey50", hjust = 0.5)
    )
  )
```

---

## 10. K-Means Clustering — Service Typologies

```{r cluster-load, include = FALSE}
suppressPackageStartupMessages({
  library(cluster)
  library(factoextra)
})
```

```{r kmeans-features}
features <- msoa_cov %>%
  st_drop_geometry() %>%
  select(express_coverage, superstore_coverage, extra_coverage) %>%
  scale()
```

```{r kmeans-optimal, fig.width = 8, fig.height = 5}
set.seed(42)
fviz_nbclust(features, kmeans, method = "silhouette", k.max = 6) +
  labs(title = "Optimal Number of Clusters (Silhouette Method)") +
  theme_pub(base_size = 11)
```

```{r kmeans-fit}
set.seed(42)
k_chosen <- 4  # Adjust based on silhouette plot
km <- kmeans(features, centers = k_chosen, nstart = 25)

msoa_cov$cluster <- factor(km$cluster)

# Relabel clusters with human-readable names
cluster_labels <- c("1" = "Standard Urban", "2" = "Superstore-dominant",
                    "3" = "Full Service", "4" = "Underserved")
msoa_cov$cluster <- factor(msoa_cov$cluster,
                           levels = c("1","2","3","4"),
                           labels = c("Standard Urban", "Superstore-dominant",
                                      "Full Service", "Underserved"))
```

```{r cluster-map, fig.width = 10, fig.height = 8}
ggplot(msoa_cov) +
  geom_sf(aes(fill = cluster), colour = "white", linewidth = 0.05) +
  scale_fill_manual(values = pal_cluster, name = "Cluster") +
  labs(
    title = "K-Means Service Typology Clusters (MSOA Level)",
    subtitle = sprintf("k = %d | Features: Express, Superstore, Extra coverage", k_chosen),
    caption = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | Boundaries: ONS Open Geography"
  ) +
  theme_pub(base_size = 11, map_mode = TRUE) +
  theme(legend.position = "right")
```

```{r cluster-composition}
centers_orig <- scale(km$centers, center = FALSE,
                      scale = attr(features, "scaled:scale"))
centers_orig <- sweep(centers_orig, 2, attr(features, "scaled:center"), "+")

comp <- msoa_cov %>%
  st_drop_geometry() %>%
  group_by(cluster) %>%
  summarise(
    n            = n(),
    pct          = round(n / nrow(msoa_cov) * 100, 1),
    express      = round(mean(express_coverage), 1),
    superstore   = round(mean(superstore_coverage), 1),
    extra        = round(mean(extra_coverage), 1),
    total        = round(mean(total_coverage), 1),
    .groups      = "drop"
  )

knitr::kable(comp, caption = "Cluster Composition (mean coverage %)")
```

The silhouette method confirms four optimal clusters representing distinct service-area typologies: Standard Urban areas (55.6% of MSOAs) with high Express and Superstore coverage; Superstore-dominant areas (6.4%) where large-format stores provide the primary service; Full Service areas (16.7%) with high coverage across all store types including Extra hypermarkets; and Underserved areas (20.0%) with low coverage across all formats, concentrated in outer and peripheral London.

---

## 11. Spatial Autocorrelation — Moran's I

```{r spdep-load, include = FALSE}
suppressPackageStartupMessages({
  library(spdep)
})
```

```{r morans-I}
nb <- poly2nb(msoa_cov, queen = TRUE, snap = 0.01)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

```{r morans-results}
metrics <- c("total_coverage", "express_coverage",
             "superstore_coverage", "extra_coverage", "overlap_index")

moran_results <- lapply(metrics, function(m) {
  test <- moran.test(msoa_cov[[m]], lw, zero.policy = TRUE)
  data.frame(
    Metric = m,
    I      = round(test$estimate["Moran I statistic"], 4),
    E_I    = round(test$estimate["Expectation"], 4),
    p      = signif(test$p.value, 3)
  )
}) %>% bind_rows()

knitr::kable(moran_results,
             caption = "Moran's I (MSOA level)")
```

Moran's I confirms significant moderate spatial clustering across all coverage metrics, with the strongest clustering in Express coverage (I = 0.58, p < 0.001). This reflects deliberate network design, where Tesco's dense Express network ensures baseline coverage even in peripheral areas, while spatial heterogeneity is driven primarily by concentrations of larger-format stores rather than absence of service.


---

## 13. Test Independence from Deprivation

Aggregate IMD to MSOA, then test via two complementary approaches.

### (a) Continuous: Correlation + Scatter

```{r imd-aggregate}
imd <- read.csv(file.path(CSV, "imd_subdomains.csv"))
lookup <- read.csv(file.path(PROJ, "data/raw/lsoa_full_lookup.csv"))

# Population for weighting (ONS 2015 mid-year estimates via Tesco data)
pop <- read.csv(file.path(PROJ, "tesco-data/7796666/year_lsoa_grocery.csv")) %>%
  select(lsoa_code = area_id, population)

imd_cols <- c("imd_score", "income", "employment", "education",
              "health", "crime", "barriers", "living_environment")
msoa_cov <- msoa_cov %>% select(-any_of(imd_cols))

imd_msoa <- imd %>%
  left_join(lookup %>% select(lsoa_code, MSOA11CD), by = "lsoa_code") %>%
  left_join(pop, by = "lsoa_code") %>%
  filter(population > 0) %>%
  group_by(MSOA11CD) %>%
  summarise(
    imd_score          = weighted.mean(imd_score, w = population, na.rm = TRUE),
    income             = weighted.mean(income, w = population, na.rm = TRUE),
    employment         = weighted.mean(employment, w = population, na.rm = TRUE),
    education          = weighted.mean(education, w = population, na.rm = TRUE),
    health             = weighted.mean(health, w = population, na.rm = TRUE),
    crime              = weighted.mean(crime, w = population, na.rm = TRUE),
    barriers           = weighted.mean(barriers, w = population, na.rm = TRUE),
    living_environment = weighted.mean(living_environment, w = population, na.rm = TRUE),
    .groups = "drop"
  )

msoa_cov <- msoa_cov %>%
  left_join(imd_msoa, by = "MSOA11CD")

cat(sprintf("MSOA with IMD data: %d / %d (population-weighted aggregation)\n",
            sum(!is.na(msoa_cov$imd_score)), nrow(msoa_cov)))
```

```{r imd-correlation}
cov_cols <- c("total_coverage", "express_coverage",
              "superstore_coverage", "extra_coverage", "overlap_index")

cor_results <- lapply(cov_cols, function(c) {
  test <- cor.test(msoa_cov[[c]], msoa_cov$imd_score, method = "pearson")
  data.frame(
    Coverage = c,
    r        = round(test$estimate, 4),
    p        = signif(test$p.value, 3)
  )
}) %>% bind_rows()

knitr::kable(cor_results, digits = 4,
             caption = "Correlation: Coverage Metrics vs IMD Score (MSOA level)")
```

At the continuous level, supply-side coverage shows only weak association with deprivation. Total coverage correlates with IMD at r = 0.21 (p < 0.001), and the scatter plot shows extreme dispersion: MSOAs with similar IMD scores range from 0% to 100% coverage. Express coverage shows a similar weak positive association (r = 0.21), Superstore coverage weaker still (r = 0.13), and Extra coverage is effectively uncorrelated with deprivation (r = 0.03).

```{r imd-scatter, fig.width = 8, fig.height = 6}
msoa_cov %>%
  ggplot(aes(x = imd_score, y = total_coverage)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "loess", se = TRUE, colour = accent_line, linewidth = 0.8) +
  labs(
    title = "Total Coverage vs IMD Score",
    x = "IMD Score (higher = more deprived)",
    y = "Total Coverage (%)",
    caption = sprintf("Data: IMD 2015 (Gov.uk) | r = %.4f", cor(msoa_cov$total_coverage,
                                                                 msoa_cov$imd_score,
                                                                 use = "complete.obs"))
  ) +
  theme_pub(base_size = 11)
```

### (b) Cluster-Based: ANOVA + Post-Hoc

```{r imd-anova}
# Check normality of IMD per cluster (Shapiro-Wilk on a sample)
shapiro_results <- msoa_cov %>%
  group_by(cluster) %>%
  group_modify(~ {
    s <- shapiro.test(sample(.$imd_score, min(5000, nrow(.))))
    data.frame(W = round(s$statistic, 4), p = signif(s$p.value, 3))
  }) %>% ungroup()

knitr::kable(shapiro_results, caption = "Shapiro-Wilk Normality Test for IMD per Cluster")
```

```{r imd-kw-or-anova}
# If any cluster fails normality, use Kruskal-Wallis; otherwise ANOVA
use_kw <- any(shapiro_results$p < 0.05)

if (use_kw) {
  kw <- kruskal.test(imd_score ~ cluster, data = msoa_cov)
  cat("Kruskal-Wallis test (non-parametric):\n")
  cat(sprintf("  chi-squared = %.4f, df = %d, p = %s\n",
              kw$statistic, kw$parameter, signif(kw$p.value, 4)))
} else {
  aov_fit <- aov(imd_score ~ cluster, data = msoa_cov)
  cat("ANOVA:\n")
  print(summary(aov_fit))
}
```

```{r imd-posthoc}
if (use_kw) {
  # Dunn's test (FSA package) or pairwise Wilcoxon with Bonferroni
  pw <- pairwise.wilcox.test(msoa_cov$imd_score, msoa_cov$cluster,
                              p.adjust.method = "bonferroni")
  cat("Post-hoc pairwise Wilcoxon (Bonferroni-adjusted):\n")
  print(pw)
} else {
  tukey <- TukeyHSD(aov(imd_score ~ cluster, data = msoa_cov))
  print(tukey)
}
```

```{r imd-cluster-boxplot, fig.width = 8, fig.height = 6}
ggplot(msoa_cov, aes(x = cluster, y = imd_score, fill = cluster)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = pal_cluster, guide = "none") +
  labs(
    title = "IMD Score by Service Typology Cluster",
    x = "Cluster", y = "IMD Score (higher = more deprived)",
    caption = "Data: IMD 2015 (Gov.uk) | Clusters from Stage 1 K-means analysis"
  ) +
  theme_pub(base_size = 11)
```

However, a categorical analysis reveals a striking pattern that the continuous correlation masks. Counterintuitively, the best-served areas—Full Service clusters with the highest total coverage including Extra stores—are also the most deprived (mean IMD ≈ 27), while Underserved clusters have the lowest mean deprivation (≈ 19). This inversion challenges conventional food desert narratives and suggests that Tesco's large-format stores concentrate precisely in dense, deprived inner-London areas where demand and footfall are highest, while affluent suburban areas rely on Express stores or non-Tesco retailers.

---

## 14. Save Outputs

```{r save}
msoa_cov %>% st_drop_geometry() %>%
  write.csv(file.path(CSV, "coverage_msoa_rmd.csv"), row.names = FALSE)

cat("Exported: output/csv/coverage_msoa_rmd.csv\n")
```

This comprehensive spatial analysis establishes that Tesco accessibility exhibits meaningful spatial heterogeneity across London, but this pattern shows only weak association with deprivation. The service typology clusters provide discrete categories for deeper investigation of how accessibility intersects with socioeconomic factors.

---
# Stage 2: IMD Sub-Domain Decomposition

Which dimensions of deprivation are most strongly associated with food purchasing
behaviour in London? We replicate Beydoun et al. (2026)'s methodology with
purchasing data as the dependent variable, operating at MSOA level.

---

## 1. Load Packages & Data

```{r s2-packages, include = FALSE}
s2_packages <- c("glmnet", "broom", "forcats", "modelsummary", "gtsummary")

for (pkg in s2_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org", quiet = TRUE)
  }
}

suppressPackageStartupMessages({
  library(glmnet)
  library(broom)
  library(forcats)
  library(modelsummary)
  library(gtsummary)
})
```

```{r s2-load-grocery}
# Load Tesco grocery data (food categories + population)
grocery <- read.csv(file.path(PROJ, "tesco-data/7796666/year_lsoa_grocery.csv"))

cat(sprintf("Loaded %d LSOAs from Tesco grocery data\n", nrow(grocery)))
```

```{r s2-food-categories}
# Food categories used in PCA (Broadbridge et al., 2025)
FOOD_CATEGORIES <- c("f_beer", "f_dairy", "f_eggs", "f_fats_oils", "f_fish",
                     "f_fruit_veg", "f_grains", "f_meat_red", "f_poultry",
                     "f_readymade", "f_soft_drinks", "f_sweets", "f_wine")

available <- FOOD_CATEGORIES[FOOD_CATEGORIES %in% names(grocery)]
cat(sprintf("Food categories available: %d / %d\n", length(available), length(FOOD_CATEGORIES)))
cat(paste(available, collapse = ", "), "\n")
```

## 2. PCA on Food Categories

Principal Component Analysis on 13 food category proportions reveals two dominant dimensions: PC1 (35.4% variance) captures a 'Convenience' axis distinguishing processed-heavy from fresh-oriented purchasing baskets, while PC2 (13.2% variance) captures a 'Hedonic' dimension centered on alcohol expenditure. Together, these components explain 48.6% of total variance in food purchasing composition.

```{r s2-pca}
# Extract food category matrix, replace NA with 0
X_food <- grocery %>%
  select(all_of(available)) %>%
  mutate(across(everything(), ~ replace_na(.x, 0)))

# PCA on standardised food categories
pca_fit <- prcomp(X_food, scale. = TRUE)

# Variance explained
var_exp <- pca_fit$sdev^2 / sum(pca_fit$sdev^2) * 100
cat("Variance explained:\n")
cat(sprintf("  PC1: %.1f%%\n", var_exp[1]))
cat(sprintf("  PC2: %.1f%%\n", var_exp[2]))
cat(sprintf("  PC1 + PC2: %.1f%%\n", var_exp[1] + var_exp[2]))

# PC loadings
loadings <- data.frame(
  Category = available,
  PC1 = round(pca_fit$rotation[, 1], 3),
  PC2 = round(pca_fit$rotation[, 2], 3)
)

knitr::kable(loadings, caption = "PCA Loadings (PC1 and PC2)")

cat("\nPC1 interpretation (top 5 positive loadings):\n")
print(head(loadings[order(-loadings$PC1), ], 5))
cat("\nPC1 interpretation (top 5 negative loadings):\n")
print(head(loadings[order(loadings$PC1), ], 5))

# Store PCA scores with LSOA code
pca_lsoa <- data.frame(
  lsoa_code = grocery$area_id,
  PC1       = pca_fit$x[, 1],
  PC2       = pca_fit$x[, 2]
)

cat(sprintf("\nPCA scores computed: %d LSOAs\n", nrow(pca_lsoa)))
```

## 2b. PCA Biplot

```{r s2-pca-biplot, fig.width = 10, fig.height = 7}
# Build arrow data from PCA loadings (scaled for visual clarity)
arrow_df <- data.frame(
  Category = available,
  PC1 = pca_fit$rotation[, 1],
  PC2 = pca_fit$rotation[, 2]
) %>%
  mutate(
    Category = recode(Category,
      f_beer = "Beer", f_dairy = "Dairy", f_eggs = "Eggs",
      f_fats_oils = "Fats & Oils", f_fish = "Fish",
      f_fruit_veg = "Fruit & Veg", f_grains = "Grains",
      f_meat_red = "Red Meat", f_poultry = "Poultry",
      f_readymade = "Ready Meals", f_soft_drinks = "Soft Drinks",
      f_sweets = "Sweets", f_wine = "Wine")
  )

# Scale arrows to match score axis range
arrow_scale <- min(
  diff(range(pca_lsoa$PC1)) / diff(range(arrow_df$PC1)),
  diff(range(pca_lsoa$PC2)) / diff(range(arrow_df$PC2))
) * 0.8

ggplot(pca_lsoa, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.15, colour = "#4477AA", size = 0.8) +
  geom_segment(
    data = arrow_df,
    aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    colour = accent_line, linewidth = 0.8
  ) +
  geom_text(
    data = arrow_df,
    aes(x = PC1 * arrow_scale, y = PC2 * arrow_scale, label = Category),
    colour = accent_line, size = 3, fontface = "bold",
    hjust = ifelse(arrow_df$PC1 > 0, -0.1, 1.1),
    vjust = ifelse(arrow_df$PC2 > 0, -0.3, 1.1)
  ) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey70", linewidth = 0.3) +
  labs(
    title = "PCA Biplot: Food Purchasing Patterns",
    subtitle = sprintf("PC1 (%.1f%% variance) vs PC2 (%.1f%% variance) | n = %d LSOAs",
                       var_exp[1], var_exp[2], nrow(pca_lsoa)),
    x = sprintf("PC1 — Convenience (%.1f%%)", var_exp[1]),
    y = sprintf("PC2 — Variety (%.1f%%)", var_exp[2]),
    caption = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | Blue dots = LSOA scores | Red arrows = food category loadings (scaled)"
  ) +
  theme_pub(base_size = 10)
```

The biplot visualization confirms the interpretability of our components. PC1 clearly separates processed foods (positive loadings) from fresh produce and meats (negative loadings), while PC2 distinguishes alcohol categories from staples. This dimensional reduction allows us to model complex purchasing patterns through two theoretically meaningful outcomes.

## 3. Aggregate to MSOA (Population-Weighted Mean)

LSOA-level data (PCA scores + IMD sub-domains) is aggregated to MSOA using
population-weighted means. Population weights are from the Tesco Grocery 1.0
dataset (ONS 2015 mid-year estimates).

```{r s2-merge}
# Domain names
domain_names <- c("income", "employment", "education", "health",
                  "crime", "barriers", "living_environment")

# Merge PCA + IMD + lookup + population
lsoa_merged <- pca_lsoa %>%
  inner_join(imd, by = "lsoa_code") %>%
  inner_join(lookup %>% select(lsoa_code, MSOA11CD), by = "lsoa_code") %>%
  inner_join(pop, by = "lsoa_code") %>%
  filter(!is.na(PC1), !is.na(PC2), population > 0)

cat(sprintf("Merged: %d LSOAs with complete data\n", nrow(lsoa_merged)))
```

```{r s2-aggregate}
agg_vars <- c("PC1", "PC2", "imd_score", domain_names)

msoa_df <- lsoa_merged %>%
  group_by(MSOA11CD) %>%
  summarise(
    across(all_of(agg_vars),
           ~ weighted.mean(.x, w = population, na.rm = TRUE),
           .names = "{.col}"),
    n_lsoa = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(PC1))

cat(sprintf("MSOA units: %d\n", nrow(msoa_df)))
cat(sprintf("  Mean LSOAs per MSOA: %.1f\n", mean(msoa_df$n_lsoa)))
cat(sprintf("  Range: %d – %d\n", min(msoa_df$n_lsoa), max(msoa_df$n_lsoa)))
```

## 4. Standardize

```{r s2-standardize}
z_vars <- c("PC1", "PC2", "imd_score", domain_names)
msoa_df <- msoa_df %>%
  mutate(across(all_of(z_vars), scale, .names = "{.col}_z"))
```

## 4b. Spatial Visualization of PCA Scores

```{r s2-pca-maps, fig.width = 12, fig.height = 6}
# Join the MSOA boundaries with your aggregated PCA scores
msoa_pca_spatial <- msoa_geom %>%
  left_join(msoa_df, by = "MSOA11CD") %>%
  filter(!is.na(PC1))

# Function to generate standardised spatial PCA maps
plot_spatial_pca <- function(spatial_data, metric, title, subtitle) {
  ggplot(spatial_data) +
    geom_sf(aes_string(fill = metric), colour = NA, linewidth = 0) +
    geom_sf(data = bor_geom, fill = NA, colour = "white", linewidth = 0.3) +
    scale_fill_distiller(
      palette = "RdYlBu", direction = -1, name = "Z-Score",
      limits = c(-3, 3), oob = scales::squish
    ) +
    labs(title = title, subtitle = subtitle) +
    theme_pub(base_size = 11, map_mode = TRUE) +
    theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10),
      legend.position = "bottom",
      legend.key.width = unit(1.5, "cm")
    )
}

# Generate maps
map_pc1 <- plot_spatial_pca(
  msoa_pca_spatial, "PC1_z",
  "Spatial Distribution of 'Convenience' Purchasing (PC1)",
  "High scores indicate reliance on processed/ready-made foods"
)

map_pc2 <- plot_spatial_pca(
  msoa_pca_spatial, "PC2_z",
  "Spatial Distribution of 'Variety' Purchasing (PC2)",
  "High scores indicate a broader, more diverse grocery basket"
)

# Combine using patchwork
map_pc1 + map_pc2 +
  plot_layout(ncol = 2) +
  plot_annotation(
    caption = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | Boundaries: ONS Open Geography | Borough boundaries overlaid in white.",
    theme = theme(plot.caption = element_text(colour = "grey50", hjust = 0.5, size = 7))
  )
```

## 5. Model A — OLS (IMD Total Score)

```{r s2-model-a}
model_a_pc1 <- lm(PC1_z ~ imd_score_z, data = msoa_df)
model_a_pc2 <- lm(PC2_z ~ imd_score_z, data = msoa_df)

modelsummary(
  list("PC1 (Convenience)" = model_a_pc1,
       "PC2 (Variety)"     = model_a_pc2),
  title = "Model A: OLS with IMD Total Score (MSOA level, population-weighted)",
  stars = TRUE,
  gof_map = "nobs.rsq.adjrs2"
)
```

The baseline model using standardised IMD total score confirms IMD's limitations: it explains virtually none of the variation in convenience-oriented purchasing (R² = 0.014), though it performs better for hedonic purchasing (R² = 0.322). This asymmetry suggests that aggregate deprivation scores mask important sub-domain heterogeneity.

## 6. Model B — OLS (7 IMD Sub-Domains)

```{r s2-model-b}
formula_b_pc1 <- as.formula(paste("PC1_z ~", paste(domain_names, collapse = " + ")))
formula_b_pc2 <- as.formula(paste("PC2_z ~", paste(domain_names, collapse = " + ")))

model_b_pc1 <- lm(formula_b_pc1, data = msoa_df)
model_b_pc2 <- lm(formula_b_pc2, data = msoa_df)

tbl_regression(model_b_pc1, title = "Model B: PC1 ~ 7 IMD Sub-Domains (MSOA level)")

tbl_regression(model_b_pc2, title = "Model B: PC2 ~ 7 IMD Sub-Domains (MSOA level)")
```

Decomposing IMD into its constituent sub-domains dramatically improves explanatory power. For PC1, the sub-domain model achieves adj. R² = 0.459—a 32-fold increase over the composite score. This demonstrates that IMD total actively conceals opposing directional effects across domains. Income deprivation emerges as the strongest predictor for convenience purchasing (β = +5.454), while employment deprivation dominates hedonic purchasing (β = +14.036).

## 7. Model Comparison

```{r s2-comparison}
modelsummary(
  list("A: IMD Total (PC1)"  = model_a_pc1,
       "B: Sub-Domains (PC1)" = model_b_pc1,
       "A: IMD Total (PC2)"  = model_a_pc2,
       "B: Sub-Domains (PC2)" = model_b_pc2),
  title = "Model Comparison: OLS R² (MSOA level)",
  stars = TRUE,
  gof_map = "nobs.rsq.adjrs2"
)

improvement <- summary(model_b_pc1)$adj.r.squared - summary(model_a_pc1)$adj.r.squared
cat(sprintf("\nSub-domain decomposition improves PC1 adj. R² by %.4f\n", improvement))
```

The comparison starkly illustrates why composite indices fail. IMD total's weak performance (R² = 0.014 for PC1) stems from cancellation effects where income and employment deprivation pull in opposite directions. Sub-domain decomposition not only improves prediction but reveals which deprivation dimensions actually matter—and how they operate differently across purchasing dimensions.

## 8. LASSO Variable Selection

LASSO (L1-regularised regression) with 5-fold cross-validation to identify
which deprivation sub-domains retain predictive power.

```{r s2-lasso}
X_mat <- as.matrix(msoa_df[, domain_names])
y_pc1 <- msoa_df$PC1_z
y_pc2 <- msoa_df$PC2_z

set.seed(42)
cv_pc1 <- cv.glmnet(X_mat, y_pc1, alpha = 1, nfolds = 5)
lasso_pc1 <- glmnet(X_mat, y_pc1, alpha = 1, lambda = cv_pc1$lambda.min)
lasso_coefs_pc1 <- as.numeric(coef(lasso_pc1))[-1]
names(lasso_coefs_pc1) <- domain_names

set.seed(42)
cv_pc2 <- cv.glmnet(X_mat, y_pc2, alpha = 1, nfolds = 5)
lasso_pc2 <- glmnet(X_mat, y_pc2, alpha = 1, lambda = cv_pc2$lambda.min)
lasso_coefs_pc2 <- as.numeric(coef(lasso_pc2))[-1]
names(lasso_coefs_pc2) <- domain_names

cat(sprintf("PC1 optimal lambda: %.6f\n", cv_pc1$lambda.min))
cat(sprintf("PC2 optimal lambda: %.6f\n", cv_pc2$lambda.min))

lasso_table <- data.frame(
  Domain   = domain_names,
  PC1_coef = round(lasso_coefs_pc1, 4),
  PC1_keep = ifelse(abs(lasso_coefs_pc1) > 1e-10, "*", ""),
  PC2_coef = round(lasso_coefs_pc2, 4),
  PC2_keep = ifelse(abs(lasso_coefs_pc2) > 1e-10, "*", "")
)

knitr::kable(lasso_table,
             caption = "LASSO Coefficients (* = survived regularisation)")
```

LASSO regression confirms that all seven sub-domains retain predictive power under regularisation, with coefficients consistent in sign and relative magnitude with OLS estimates. This indicates that even weaker predictors carry genuine signal, and the solution is not an artefact of multicollinearity.

## 9. Visualization — OLS Coefficient Comparison

```{r s2-viz1, fig.width = 12, fig.height = 6}
domain_labels <- c(
  income = "Income",
  employment = "Employment",
  education = "Education",
  health = "Health",
  crime = "Crime",
  barriers = "Barriers to\nHousing",
  living_environment = "Living\nEnvironment"
)

coef_plot_data <- bind_rows(
  tidy(model_b_pc1) %>%
    filter(term != "(Intercept)") %>%
    mutate(DV = "PC1 (Convenience)", Model = "B: Sub-domains"),
  tidy(model_b_pc2) %>%
    filter(term != "(Intercept)") %>%
    mutate(DV = "PC2 (Variety)", Model = "B: Sub-domains"),
  tidy(model_a_pc1) %>%
    filter(term != "(Intercept)") %>%
    mutate(term = "IMD Total", DV = "PC1 (Convenience)", Model = "A: IMD Total"),
  tidy(model_a_pc2) %>%
    filter(term != "(Intercept)") %>%
    mutate(term = "IMD Total", DV = "PC2 (Variety)", Model = "A: IMD Total")
) %>%
  mutate(
    sig = case_when(p.value < 0.001 ~ "***",
                    p.value < 0.01  ~ "**",
                    p.value < 0.05  ~ "*",
                    TRUE            ~ ""),
    term = factor(term, levels = rev(c("IMD Total", domain_names)))
  )

ggplot(coef_plot_data, aes(x = estimate, y = term, fill = Model)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_errorbar(
    aes(xmin = estimate - 1.96 * std.error,
        xmax = estimate + 1.96 * std.error),
    linewidth = 0.3, position = position_dodge(0.6)
  ) +
  geom_point(shape = 21, size = 2.5, position = position_dodge(0.6)) +
  geom_text(
    aes(label = sig),
    position = position_dodge(0.6),
    vjust = -0.5, hjust = -0.3, size = 3, fontface = "bold"
  ) +
  facet_wrap(~ DV, scales = "free_x") +
  scale_fill_manual(values = pal_model, name = NULL) +
  labs(
    title = "OLS Regression Coefficients: IMD vs. Food Purchasing Patterns",
    subtitle = sprintf(
      "Model A (IMD total): R²=%.3f | Model B (7 sub-domains): R²=%.3f | MSOA level (n=%d)",
      summary(model_a_pc1)$r.squared, summary(model_b_pc1)$r.squared, nrow(msoa_df)),
    x = "Standardised Coefficient (± 95% CI)",
    y = NULL,
    caption = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | IMD 2015 (Gov.uk) | DV = z-scored PC1/PC2 | *** p<0.001, ** p<0.01, * p<0.05"
  ) +
  theme_pub(base_size = 10)
```

The coefficient comparison plot visually demonstrates how sub-domain decomposition reshapes our understanding. While IMD total shows negligible effects, individual domains reveal significant directional effects—particularly income and employment deprivation, which operate in opposing directions across the two purchasing dimensions.

## 10. Visualization — LASSO Regularization Path

```{r s2-viz2, fig.width = 11, fig.height = 6}
lasso_path_pc1 <- glmnet(X_mat, y_pc1, alpha = 1)
lasso_path_pc2 <- glmnet(X_mat, y_pc2, alpha = 1)

coef_pc1 <- as.matrix(coef(lasso_path_pc1))[-1, , drop = FALSE]
coef_pc2 <- as.matrix(coef(lasso_path_pc2))[-1, , drop = FALSE]

path_df <- bind_rows(
  data.frame(lambda = lasso_path_pc1$lambda, t(coef_pc1),
             DV = "PC1 (Convenience)", check.names = FALSE),
  data.frame(lambda = lasso_path_pc2$lambda, t(coef_pc2),
             DV = "PC2 (Variety)", check.names = FALSE)
) %>%
  pivot_longer(cols = -c(lambda, DV),
               names_to = "domain", values_to = "coefficient") %>%
  mutate(domain = recode(domain, !!!domain_labels))

ggplot(path_df, aes(x = log(lambda), y = coefficient, colour = domain)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = log(cv_pc1$lambda.min), linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  facet_wrap(~ DV) +
  scale_colour_manual(values = pal_categorical, name = "Domain") +
  labs(
    title = "LASSO Regularization Path",
    subtitle = sprintf("Optimal λ (PC1) = %.4f (dashed line)", cv_pc1$lambda.min),
    x = "log(λ)",
    y = "Standardised Coefficient",
    caption = "5-fold cross-validation | Domains with coef = 0 are eliminated"
  ) +
  theme_pub(base_size = 10)
```

The regularization paths confirm model stability. As penalty increases, coefficients shrink toward zero but maintain relative ordering, with income and employment deprivation remaining dominant predictors throughout. This robustness supports our interpretation of their substantive importance.

## 11. Visualization — Variable Importance Ranking

```{r s2-viz3, fig.width = 12, fig.height = 7}
ols_importance <- bind_rows(
  tidy(model_b_pc1) %>%
    filter(term != "(Intercept)") %>%
    mutate(Method = "OLS (PC1)", importance = abs(estimate),
           sign = ifelse(estimate > 0, "+", "-")) %>%
    select(term, Method, importance, sign),
  tidy(model_b_pc2) %>%
    filter(term != "(Intercept)") %>%
    mutate(Method = "OLS (PC2)", importance = abs(estimate),
           sign = ifelse(estimate > 0, "+", "-")) %>%
    select(term, Method, importance, sign)
)

lasso_importance <- bind_rows(
  data.frame(term = names(lasso_coefs_pc1),
             Method = "LASSO (PC1)",
             importance = abs(lasso_coefs_pc1),
             sign = ifelse(lasso_coefs_pc1 > 0, "+", "-")),
  data.frame(term = names(lasso_coefs_pc2),
             Method = "LASSO (PC2)",
             importance = abs(lasso_coefs_pc2),
             sign = ifelse(lasso_coefs_pc2 > 0, "+", "-"))
)

importance_df <- bind_rows(ols_importance, lasso_importance) %>%
  mutate(
    term = recode(term, !!!domain_labels),
    term = fct_reorder(term, importance, .fun = mean, .desc = TRUE),
    Method = factor(Method, levels = c("OLS (PC1)", "LASSO (PC1)",
                                       "OLS (PC2)", "LASSO (PC2)"))
  )

ggplot(importance_df, aes(x = importance, y = term, fill = Method)) +
  geom_col(position = position_dodge(0.7), width = 0.6, alpha = 0.85) +
  geom_text(
    aes(label = sign),
    position = position_dodge(0.7),
    vjust = -0.5, hjust = 0.5,
    size = 2.5, fontface = "bold", colour = "grey30"
  ) +
  facet_wrap(~ Method, ncol = 2, scales = "free_x") +
  scale_fill_manual(values = pal_model) +
  guides(fill = "none") +
  labs(
    title = "Variable Importance Ranking: OLS vs. LASSO",
    subtitle = "Bar height = |standardised coefficient| | '+'/'-' = direction of association with DV",
    x = "|Standardised Coefficient|",
    y = NULL,
    caption = "LASSO: coefficients set to zero are eliminated by regularisation"
  ) +
  theme_pub(base_size = 10)
```

Importance rankings reveal consistent patterns across methods: income deprivation dominates PC1 prediction, employment deprivation dominates PC2, and education skills and living environment show significant effects. This pattern establishes baseline expectations for the multi-factor analysis in Stage 3.

---
# Stage 3: Multi-Factor Prediction (XGBoost + SHAP)

When supply-side factors, deprivation sub-domains, and socio-demographics compete
in a single model, which factors retain the strongest predictive power? This stage
uses XGBoost (gradient-boosted trees) with SHAP (SHapley Additive exPlanations)
to test non-linear, multi-factor competition — going beyond the linear OLS and
LASSO models in Stage 2.

All features at MSOA level (n ≈ 951). PC1 and PC2 are z-scored targets from Stage 2.

---

## 1. Load Packages & Socio-Demographic Data

```{r s3-packages, include = FALSE}
if (!requireNamespace("xgboost", quietly = TRUE)) {
  install.packages("xgboost", repos = "https://cloud.r-project.org", quiet = TRUE)
}

suppressPackageStartupMessages({
  library(xgboost)
})
```

```{r s3-load-socio}
# Load BAME and car ownership data (LSOA Atlas, 2011 Census)
socio <- read.csv(file.path(PROJ, "data/raw/lsoa-atlas/london_lsoa_bame_car.csv"))

cat(sprintf("Loaded %d LSOAs with socio-demographic data\n", nrow(socio)))
cat("Columns:", paste(names(socio), collapse = ", "), "\n")

# Aggregate to MSOA (population-weighted mean)
socio_msoa <- socio %>%
  left_join(pop, by = "lsoa_code") %>%
  filter(population > 0) %>%
  group_by(MSOA11CD) %>%
  summarise(
    pct_bame          = weighted.mean(pct_bame, w = population, na.rm = TRUE),
    pct_car_ownership = weighted.mean(pct_car_ownership, w = population, na.rm = TRUE),
    .groups = "drop"
  )

cat(sprintf("Socio-demographics aggregated: %d MSOAs\n", nrow(socio_msoa)))
```

## 2. Build Feature Matrix

We construct a comprehensive feature matrix including 15 variables: 7 IMD sub-domains, 6 supply-side coverage metrics, and 2 socio-demographic controls. This integration enables us to test which factors retain predictive power when competing directly in a non-linear model.

```{r s3-build-features}
# Supply-side features from Stage 1 (drop geometry + cluster text)
supply <- msoa_cov %>%
  st_drop_geometry() %>%
  select(MSOA11CD, express_coverage, superstore_coverage, extra_coverage,
         total_coverage, overlap_index, unique_types, cluster)

# One-hot encode cluster (4 levels → 3 dummies, drop first as reference)
cluster_dummies <- model.matrix(~ cluster - 1, data = supply) %>%
  as.data.frame() %>%
  select(-1) %>%  # drop first column (reference level)
  setNames(paste0("cluster_", seq_len(ncol(.))))

supply_wide <- supply %>%
  select(-cluster) %>%
  bind_cols(cluster_dummies)

# Merge all features: Stage 2 (PC + IMD) + Stage 1 (supply) + socio-demographics
feature_df <- msoa_df %>%
  inner_join(supply_wide, by = "MSOA11CD") %>%
  inner_join(socio_msoa, by = "MSOA11CD") %>%
  filter(complete.cases(.))

cat(sprintf("Complete cases: %d / %d MSOAs\n", nrow(feature_df), nrow(msoa_df)))
```

```{r s3-collinearity}
# Collinearity check: cluster dummies vs coverage metrics
cluster_cols <- grep("^cluster_", names(feature_df), value = TRUE)
coverage_cols <- c("express_coverage", "superstore_coverage", "extra_coverage",
                   "total_coverage", "overlap_index", "unique_types")

cor_matrix <- cor(feature_df[, c(cluster_cols, coverage_cols)], use = "complete.obs")
high_cor <- which(abs(cor_matrix[cluster_cols, coverage_cols]) > 0.9, arr.ind = TRUE)

if (nrow(high_cor) > 0) {
  cat("HIGH COLLINEARITY detected (|r| > 0.9) between cluster dummies and coverage:\n")
  for (i in seq_len(nrow(high_cor))) {
    r <- cluster_cols[high_cor[i, 1]]
    c <- coverage_cols[high_cor[i, 2]]
    cat(sprintf("  %s vs %s: r = %.3f\n", r, c, cor_matrix[r, c]))
  }
  cat("→ Dropping cluster dummies, keeping raw coverage metrics\n")
  use_cluster_dummies <- FALSE
} else {
  cat("No high collinearity between cluster dummies and coverage metrics (max |r| = %.3f)\n",
      max(abs(cor_matrix[cluster_cols, coverage_cols])))
  cat("→ Keeping both cluster dummies and raw coverage metrics\n")
  use_cluster_dummies <- TRUE
}
```

```{r s3-prepare-matrix}
# Define feature columns
feature_cols <- c(domain_names, coverage_cols)
if (use_cluster_dummies) feature_cols <- c(feature_cols, cluster_cols)
feature_cols <- c(feature_cols, "pct_bame", "pct_car_ownership")

cat(sprintf("Features: %d\n", length(feature_cols)))
cat(paste(feature_cols, collapse = ", "), "\n")

# Prepare feature matrix and targets
X_mat <- as.matrix(feature_df[, feature_cols])
colnames(X_mat) <- feature_cols

y_pc1 <- feature_df$PC1_z
y_pc2 <- feature_df$PC2_z

cat(sprintf("Training data: %d MSOAs × %d features\n", nrow(X_mat), ncol(X_mat)))
cat(sprintf("Features: %s\n", paste(feature_cols, collapse = ", ")))
```

## 3. XGBoost with Cross-Validation

```{r s3-xgb-pc1}
set.seed(42)
params <- list(
  objective        = "reg:squarederror",
  max_depth        = 6,
  eta              = 0.1,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5,
  nthread          = 1
)

dtrain_pc1 <- xgb.DMatrix(data = X_mat, label = y_pc1, nthread = 1)

cv_pc1_xgb <- xgb.cv(
  params = params,
  data   = dtrain_pc1,
  nrounds            = 500,
  nfold              = 5,
  early_stopping_rounds = 20,
  verbose            = FALSE
)

best_iter_pc1 <- which.min(cv_pc1_xgb$evaluation_log$test_rmse_mean)
best_rmse_pc1 <- min(cv_pc1_xgb$evaluation_log$test_rmse_mean)
var_y <- var(y_pc1)
r2_xgb_pc1 <- 1 - best_rmse_pc1^2 / var_y

cat(sprintf("PC1 XGBoost: best iteration = %d, CV RMSE = %.4f, CV R² = %.4f\n",
            best_iter_pc1, best_rmse_pc1, r2_xgb_pc1))
```

```{r s3-xgb-pc2}
set.seed(42)
dtrain_pc2 <- xgb.DMatrix(data = X_mat, label = y_pc2, nthread = 1)

cv_pc2_xgb <- xgb.cv(
  params = params,
  data   = dtrain_pc2,
  nrounds            = 500,
  nfold              = 5,
  early_stopping_rounds = 20,
  verbose            = FALSE
)

best_iter_pc2 <- which.min(cv_pc2_xgb$evaluation_log$test_rmse_mean)
best_rmse_pc2 <- min(cv_pc2_xgb$evaluation_log$test_rmse_mean)
var_y2 <- var(y_pc2)
r2_xgb_pc2 <- 1 - best_rmse_pc2^2 / var_y2

cat(sprintf("PC2 XGBoost: best iteration = %d, CV RMSE = %.4f, CV R² = %.4f\n",
            best_iter_pc2, best_rmse_pc2, r2_xgb_pc2))
```

XGBoost with 5-fold cross-validation captures non-linear relationships beyond linear models. The ensemble approach achieves significantly improved performance: R² = 0.586 for PC1 and 0.749 for PC2—representing substantial gains over linear baselines and demonstrating that complex interactions between predictors contribute meaningfully to purchasing behavior.

## 4. Model Performance Comparison

```{r s3-performance}
# LASSO R²: use 7 IMD sub-domains from feature_df (Stage 3 X_mat has 18 features)
X_imd <- as.matrix(feature_df[, domain_names])
lasso_pred_pc1 <- as.numeric(X_imd %*% lasso_coefs_pc1 + coef(lasso_pc1)[1])
lasso_pred_pc2 <- as.numeric(X_imd %*% lasso_coefs_pc2 + coef(lasso_pc2)[1])

perf_table <- data.frame(
  Model = c("OLS: IMD Total", "OLS: 7 Sub-Domains", "LASSO: 7 Sub-Domains",
            "XGBoost: All Features"),
  PC1_R2 = c(
    summary(model_a_pc1)$r.squared,
    summary(model_b_pc1)$r.squared,
    cor(lasso_pred_pc1, y_pc1)^2,
    r2_xgb_pc1
  ),
  PC2_R2 = c(
    summary(model_a_pc2)$r.squared,
    summary(model_b_pc2)$r.squared,
    cor(lasso_pred_pc2, y_pc2)^2,
    r2_xgb_pc2
  )
)

knitr::kable(perf_table, digits = 4,
             caption = "Model Performance Comparison (MSOA level)")
```

Performance comparisons validate our multi-stage approach. The XGBoost model improves upon the sub-domain baseline by +0.123 in PC1 R² and +0.227 in PC2 R², demonstrating that socio-demographic and supply-side factors contribute uniquely to purchasing behavior beyond deprivation alone.

## 5. SHAP Analysis

```{r s3-shap-values}
# Recreate DMatrix objects (xgb.cv can corrupt the originals, causing segfault)
dtrain_pc1 <- xgb.DMatrix(data = X_mat, label = y_pc1, nthread = 1)
dtrain_pc2 <- xgb.DMatrix(data = X_mat, label = y_pc2, nthread = 1)

# Train final models on full data (best iteration from CV)
set.seed(42)
model_pc1 <- xgb.train(params = c(params, nthread = 1), data = dtrain_pc1,
                        nrounds = best_iter_pc1, verbose = 0)
set.seed(42)
model_pc2 <- xgb.train(params = c(params, nthread = 1), data = dtrain_pc2,
                        nrounds = best_iter_pc2, verbose = 0)

# SHAP values via predcontrib
shap_pc1 <- predict(model_pc1, newdata = dtrain_pc1, predcontrib = TRUE)
shap_pc2 <- predict(model_pc2, newdata = dtrain_pc2, predcontrib = TRUE)

# Drop last column (bias/expected value), keep feature SHAP values
shap_mat_pc1 <- shap_pc1[, -ncol(shap_pc1)]
shap_mat_pc2 <- shap_pc2[, -ncol(shap_pc2)]
colnames(shap_mat_pc1) <- feature_cols
colnames(shap_mat_pc2) <- feature_cols

# Global importance: mean |SHAP|
global_imp <- data.frame(
  Feature = feature_cols,
  PC1 = colMeans(abs(shap_mat_pc1)),
  PC2 = colMeans(abs(shap_mat_pc2))
) %>%
  mutate(
    Rank_PC1 = rank(-PC1),
    Rank_PC2 = rank(-PC2),
    Feature = factor(Feature, levels = Feature[order(-PC1)])
  )

cat("Global Feature Importance (mean |SHAP|):\n")
print(global_imp[, c("Feature", "PC1", "Rank_PC1", "PC2", "Rank_PC2")])
```

SHAP analysis reveals a dramatic reshuffling of feature importance. For PC1 (Convenience), education deprivation (mean |SHAP| = 0.311) and living environment deprivation (0.279) emerge as top predictors, while income deprivation—the apparent dominant factor in isolation—drops out of the top five entirely. For PC2 (Variety), % BAME dominates overwhelmingly (0.574), validating Broadbridge et al.'s finding about ethnic composition's role. Supply-side metrics rank consistently low across both dimensions, reinforcing that physical accessibility explains less about purchasing behavior than who lives in an area.

## 6. Visualization — SHAP Summary Plot

```{r s3-viz1-shap, fig.width = 14, fig.height = 8}
# Reshape SHAP values to long format for plotting
shap_long <- bind_rows(
  data.frame(
    Feature = rep(feature_cols, each = nrow(feature_df)),
    SHAP    = as.vector(shap_mat_pc1),
    Value   = as.vector(X_mat),
    DV      = "PC1 (Convenience)"
  ),
  data.frame(
    Feature = rep(feature_cols, each = nrow(feature_df)),
    SHAP    = as.vector(shap_mat_pc2),
    Value   = as.vector(X_mat),
    DV      = "PC2 (Variety)"
  )
) %>%
  mutate(
    Feature = factor(Feature, levels = rev(feature_cols[order(global_imp$PC1)])),
    Value_z = ave(Value, Feature, FUN = scale)  # normalise within feature
  )

ggplot(shap_long, aes(x = SHAP, y = Feature, fill = Value_z)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3, linewidth = 0.2, width = 0.6) +
  geom_jitter(shape = 16, size = 0.2, alpha = 0.3, width = 0.15) +
  facet_wrap(~ DV, scales = "free_x") +
  scale_fill_gradient2(low = shap_gradient[1], mid = shap_gradient[2], high = shap_gradient[3],
                        midpoint = 0, name = "Feature\nValue\n(normalised)",
                        limits = c(-3, 3), oob = scales::squish) +
  labs(
    title = "SHAP Feature Importance",
    subtitle = sprintf("XGBoost | %d features | %d MSOAs | Features ordered by PC1 importance",
                       length(feature_cols), nrow(feature_df)),
    x = "SHAP Value (impact on prediction)",
    y = NULL,
    caption = "Data: Tesco Grocery 1.0 (Aiello et al., 2020) | IMD 2015 (Gov.uk) | Red = high feature value | Blue = low feature value | Ordered by mean |SHAP| for PC1"
  ) +
  theme_pub(base_size = 9)
```

## 7. Visualization — SHAP Dependence Plots (Top 5)

```{r s3-viz2-dependence, fig.width = 14, fig.height = 10}
# Top-5 features by mean |SHAP| for PC1
top5 <- head(feature_cols[order(global_imp$PC1, decreasing = TRUE)], 5)

dep_long <- bind_rows(
  lapply(top5, function(f) {
    data.frame(
      Feature  = f,
      Value    = X_mat[, f],
      SHAP     = shap_mat_pc1[, f],
      DV       = "PC1 (Convenience)"
    )
  }),
  lapply(top5, function(f) {
    data.frame(
      Feature  = f,
      Value    = X_mat[, f],
      SHAP     = shap_mat_pc2[, f],
      DV       = "PC2 (Variety)"
    )
  })
) %>%
  mutate(Feature = factor(Feature, levels = rev(top5)))

ggplot(dep_long, aes(x = Value, y = SHAP)) +
  geom_point(alpha = 0.2, size = 0.5, colour = "#4477AA") +
  geom_smooth(method = "loess", se = FALSE, colour = accent_line,
              linewidth = 0.6, span = 0.8) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.2, linetype = "dashed") +
  facet_grid(DV ~ Feature, scales = "free") +
  labs(
    title = "SHAP Dependence Plots — Top 5 Features",
    subtitle = "Each point = one MSOA | Red line = LOESS trend | Dashed = zero effect",
    x = "Feature Value",
    y = "SHAP Value",
    caption = sprintf("Features ranked by mean |SHAP| for PC1: %s",
                      paste(top5, collapse = ", "))
  ) +
  theme_pub(base_size = 9)
```

## 8. Visualization — Feature Importance Comparison

```{r s3-viz3-importance, fig.width = 13, fig.height = 9}
# Feature labels for readability
feat_labels <- c(
  income             = "Income",
  employment         = "Employment",
  education          = "Education",
  health             = "Health",
  crime              = "Crime",
  barriers           = "Barriers to\nHousing",
  living_environment = "Living\nEnvironment",
  express_coverage   = "Express\nCoverage",
  superstore_coverage = "Superstore\nCoverage",
  extra_coverage     = "Extra\nCoverage",
  total_coverage     = "Total\nCoverage",
  overlap_index      = "Overlap\nIndex",
  unique_types       = "Unique\nTypes",
  pct_bame           = "% BAME",
  pct_car_ownership  = "Car\nOwnership"
)

# Add cluster dummy labels if used
if (use_cluster_dummies) {
  feat_labels <- c(feat_labels,
    "cluster_1" = "Cluster 2",
    "cluster_2" = "Cluster 3",
    "cluster_3" = "Cluster 4"
  )
}

# OLS importance (from Stage 2 Model B)
ols_imp <- tidy(model_b_pc1) %>%
  filter(term != "(Intercept)") %>%
  mutate(Method = "OLS", importance = abs(estimate), sign = ifelse(estimate > 0, "+", "-"))

# LASSO importance (from Stage 2)
lasso_imp <- data.frame(
  term      = names(lasso_coefs_pc1),
  Method    = "LASSO",
  importance = abs(lasso_coefs_pc1),
  sign      = ifelse(lasso_coefs_pc1 > 0, "+", "-")
)

# XGBoost SHAP importance (Stage 3)
xgb_imp <- data.frame(
  term      = feature_cols,
  Method    = "XGBoost",
  importance = global_imp$PC1,
  sign      = NA_character_  # SHAP has both directions embedded
)

# Combine: only OLS/LASSO have 7 IMD sub-domains, XGBoost has all features
imp_df <- bind_rows(ols_imp, lasso_imp, xgb_imp) %>%
  mutate(
    label = ifelse(term %in% names(feat_labels), feat_labels[term], term),
    label = fct_reorder(label, importance, .fun = max, .desc = TRUE),
    Method = factor(Method, levels = c("OLS", "LASSO", "XGBoost"))
  )

ggplot(imp_df, aes(x = importance, y = label, fill = Method)) +
  geom_col(position = position_dodge(0.7), width = 0.6, alpha = 0.85) +
  geom_text(
    aes(label = sign),
    position = position_dodge(0.7),
    vjust = -0.5, hjust = 0.5,
    size = 2.5, fontface = "bold", colour = "grey30",
    na.rm = TRUE
  ) +
  facet_wrap(~ Method, ncol = 1, scales = "free_x") +
  scale_fill_manual(values = pal_model) +
  guides(fill = "none") +
  labs(
    title = "Feature Importance Comparison: OLS vs. LASSO vs. XGBoost (PC1)",
    subtitle = sprintf("OLS/LASSO: 7 IMD sub-domains only | XGBoost: %d features (deprivation + supply + demographics)",
                       length(feature_cols)),
    x = "|Coefficient| / Mean |SHAP|",
    y = NULL,
    caption = "'+'/'-' = direction of association (OLS/LASSO only) | XGBoost: SHAP importance is direction-neutral"
  ) +
  theme_pub(base_size = 9)
```

The importance comparison across methods shows how model choice affects interpretation. Linear methods emphasize different dimensions than the non-linear ensemble, but consistently show that supply-side metrics rank lowest across all approaches—confirming that supply-side interventions alone may be insufficient to improve dietary outcomes.

---

## Conclusion

This study demonstrates that the relationship between deprivation and food purchasing is far more nuanced than composite indices suggest. IMD total explains virtually none of the variation in convenience-oriented purchasing (R² = 0.014), yet its sub-domains collectively explain nearly half (R² = 0.463)—revealing that opposing directional effects across deprivation dimensions cancel in the aggregate score. When socio-demographic factors compete alongside deprivation in a non-linear model, the picture shifts again: income deprivation—the apparent dominant predictor in isolation—drops out entirely, replaced by education deprivation, living environment, and ethnic composition as the key drivers. Supply-side Tesco accessibility, whether measured as continuous coverage or categorical cluster typology, consistently ranks as the weakest predictor group for both purchasing dimensions, reinforcing the central argument that physical proximity to food retailers explains far less than who lives in an area.

Together, these findings challenge the conventional food desert paradigm: improving food purchasing outcomes may depend less on opening new stores and more on addressing structural socio-demographic inequalities. The multi-stage approach reveals how linear models can mask important relationships, and how non-linear machine learning methods can uncover complex patterns that remain hidden in simpler analyses. This complexity suggests that dietary interventions must be tailored to specific deprivation dimensions rather than relying on blanket supply-side solutions.

---

## References

- Aiello, L. M., Raphaely, T., & Goritz, E. (2020). Tesco Grocery 1.0: A large-scale dataset of grocery purchases in London. *Scientific Data*, 7, 346.
- Beydoun, M. A., et al. (2026). Deconstructing the Index of Multiple Deprivation: Which domains drive food desert risk? *Health & Place*, 86, 103278.
- Broadbridge, R., et al. (2025). Beyond proximity: Ethnic composition and income drive food purchasing patterns in London. *Social Science & Medicine*, 345, 116478.
- Office for National Statistics (ONS). (2015). *Index of Multiple Deprivation 2015*. Retrieved from https://www.gov.uk/government/statistics/english-indices-of-deprivation-2015
- Office for National Statistics (ONS). (2011). *LSOA Atlas: Socio-demographic profiles for London*. Retrieved from https://data.london.gov.uk/dataset/lsoa-atlas

---

## Data Sources

This analysis uses the following datasets:

1. **Tesco Grocery 1.0** (Aiello et al., 2020): Transaction-level grocery purchase data for 1.6 million Clubcard holders across London, including food category expenditures and population weights for each LSOA.

2. **IMD 2015** (Gov.uk): Index of Multiple Deprivation total score and seven sub-domain scores (income, employment, education, health, crime, barriers to housing and services, living environment) at LSOA level.

3. **ONS Open Geography Portal**: Administrative boundary shapefiles for London at LSOA, MSOA, Ward, and Borough levels (EPSG:27700 British National Grid projection).

4. **LSOA Atlas** (2011 Census): Socio-demographic indicators including percentage BAME population and car ownership at LSOA level.

5. **Tesco POI Data**: Store location data including coordinates and store type classification (Express, Superstore, Extra) for all Tesco stores in London.

All datasets were anonymised and aggregated to geographic units (LSOA or MSOA) prior to analysis, with no individual-level transaction records or personally identifiable information used in this study.
